Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-05-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Clean up manuscript before feedback

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. TODO: add stratification. \\
\hline
\end{tabular}
\end{table}

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:6, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "", "N", rep("", 30)),
        miss = c(rep("", 14), "NA", "NA", "", "", "NA", "", "", "NA", rep("", 14)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:6, labels = c(1:4, "...", "n"), name = NULL, trans = "reverse") +
  coord_cartesian(expand = c(0,0,0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: add fifth case with cluster 3
# TODO: use math font
```
Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.

A key characteristic of the missing data to take into account in analyses is the mechanism behind the missingness. We distinguish between MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice this distinction is less clear. Since the essence of the true non-response mechanism may not be known, it is generally inferred or assumed to be ignorable (i.e., MCAR or MAR). [TODO: add that this assumption may not always be valid, especially with modern types of big dasta sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.

@@

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset"). Add that the more the random effects are of interest, the more you need ml models. And add overview of possible predictor matrix values.]

@@

**TODO: add where novice and more experienced readers should start reading.** For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

@@

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(broom.mixed)  # for multilevel estimates
library(mitml)        # for multilevel pooling
```

@@

est_true <- lme4::lmer(mod, popcomp) %>% broom.mixed::tidy()

@@

This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
 **TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?**

@@

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
```

**TODO: remove the broom.mixed output, use mitml only**

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
fit_ignored <- imp_ignored %>% 
  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
```

@@

**TODO: Add that it doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low.**

@@

# impute the data, cluster as predictor
imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
```

@@

The `impact` data included in the \pkg{metamisc} package is a complete data set. The original data has already been imputed once (Steyerberg et al, 2008). For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation). The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com). TODO: email script to thomas.

@@

We will use the following estimates as comparative truth in this tutorial [TODO: make this a table or forest plot instead, to see if there is heterogeneity in the association of `ct` with `mort`]:
```{r truth}
fit <- glmer(mod, family = "binomial", data = impact) # fit the model
tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

```{r forest, echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(impact$name),
    ~ {
      glm(
        mort ~ 1 + age + motor_score + pupil + ct,
        family = "binomial",
        data = filter(impact, name == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(name = as.character(.x), .)
  )
filter(fits, term %in% c("ctIII")) %>% #, "ctIV/V"
  ggplot(aes(y = name, x = estimate)) +
  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
  geom_point() +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic()
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```

@@

This shows... [TODO: fill in that we need to impute `ct` and `pupil`.] TODO: remove axis labels

@@

The first imputation model that we'll use is likely to be invalid. We do *not* use the cluster identifier `name` as imputation model predictor. With this model, we ignore the multilevel structure of the data, despite the high ICC. This assumes exchangeability between units. We include it purely to illustrate the effects of ignoring the clustering in our imputation effort. We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).

---

\begin{center}

Updated until here!

\end{center}

---

>>>>

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
\hline
\end{tabular}
\end{table}

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  coord_cartesian(expand = c(0,0,0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.

A key characteristic of the missing data to take into account in analyses is the mechanism behind the missingness. We distinguish between MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice this distinction is less clear. Since the essence of the true non-response mechanism may not be known, it is generally inferred or assumed to be ignorable (i.e., MCAR or MAR). [TODO: add that this assumption may not always be valid, especially with modern types of big data sources.] Depending on the actual missingness-generating mechanism, missing data handling strategies may be more or less suitable, see e.g., @yuce08 and @hox15.

@@

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.]

@@

[TODO: add where novice vs more experienced readers should start reading.] For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

@@

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel pooling
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

est_true <- lme4::lmer(mod, popcomp) %>% broom.mixed::tidy()
# testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)

@@

This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).
 [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?]

@@

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

<!-- [TODO: remove the broom.mixed output, use mitml only] -->

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
```

Print the estimates:
```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

[TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]

@@

# impute the data, cluster as predictor
imp <- mice(popmis, pred = pred, print = FALSE)
```

Analyze the imputations:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
```

Print the estimates:
```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

The `impact` data included in the \pkg{metamisc} package is a complete data set. The original data has already been imputed once (Steyerberg et al, 2008). For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation). The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).

@@

The estimated effects in the complete data are visualized in Figure \ref{}.

```{r forest, echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(impact$name),
    ~ {
      glm(
        mort ~ 1 + age + motor_score + pupil + ct,
        family = "binomial",
        data = filter(impact, name == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(name = as.character(.x), .)
  )
filter(fits, term %in% c("ctIII")) %>% #, "ctIV/V"
  ggplot(aes(y = name, x = estimate)) +
  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
  geom_point() +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic()
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```
<!-- We will use the following estimates as comparative truth in this tutorial: -->
```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

@@

This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]

@@

<!-- The first imputation model that we'll use is likely to be invalid. We do *not* use the cluster identifier `name` as imputation model predictor. With this model, we ignore the multilevel structure of the data, despite the high ICC. This assumes exchangeability between units. We include it purely to illustrate the effects of ignoring the clustering in our imputation effort. We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).  -->

<!-- --- -->

<!-- \begin{center} -->

<!-- Updated until here! -->

<!-- \end{center} -->

<!-- --- -->

### 2022-05-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Clean up manuscript before feedback

A key characteristic of the missing data to take into account in
analyses is the mechanism behind the missingness. We distinguish between
MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice
this distinction is less clear. Since the essence of the true
non-response mechanism may not be known, it is generally inferred or
assumed to be ignorable (i.e., MCAR or MAR). {[}TODO: add that this
assumption may not always be valid, especially with modern types of big
dasta sources.{]} Depending on the actual missingness-generating
mechanism, missing data handling strategies may be more or less
suitable, see e.g., \citet{yuce08} and \citet{hox15}.

@@

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset''). Add that the more the
random effects are of interest, the more you need ml models. And add
overview of possible predictor matrix values.{]}

@@

\textbf{TODO: add where novice and more experienced readers should start
reading.} For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(broom.mixed)  # for multilevel estimates
R> library(mitml)        # for multilevel pooling
\end{CodeInput}
\end{CodeChunk}

@@

This shows us that there are no apparent differences in the distribution
of \texttt{texp} depending on the missingness indicator of
\texttt{popular} (t = -0.873, p = 0.383). \textbf{TODO: think about what
is a meaningful rule of thumb to signal that the user should be
worried?}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

\textbf{TODO: remove the broom.mixed output, use mitml only}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit_ignored <- imp_ignored %>% 
+   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
R> testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
\end{CodeInput}
\begin{CodeOutput}

@@

testEstimates(model = as.mitml.result(fit_ignored), var.comp = TRUE)

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     3.489     0.179    19.542     9.227     0.000     1.927     0.714 
sex             1.138     0.197     5.776    11.588     0.000     1.424     0.644 
texp            0.100     0.014     7.339     6.914     0.000     3.177     0.809 
sex:texp       -0.019     0.015    -1.217     7.708     0.260     2.576     0.773

                            Estimate 
Intercept~~Intercept|school    0.123 
sex~~sex|school                0.143 
Intercept~~sex|school          0.014 
Residual~~Residual             0.656 
ICC|school                     0.158

@@

\textbf{TODO: Add that it doesn't work with syst missing (only
sporadically). There's some pro's and con's. May not differ much if the
number of clusters is low.}

@@

\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

The \texttt{impact} data included in the \pkg{metamisc} package is a
complete data set. The original data has already been imputed once
(Steyerberg et al, 2008). For the purpose of this tutorial we have
induced missingness (mimicking the missing data in the original data set
before imputation). The resulting incomplete data can be accessed from
\href{https://zenodo.com}{zenodo link to be created}. TODO: email script
to thomas.

@@

We will use the following estimates as comparative truth in this
tutorial {[}TODO: make this a table or forest plot instead, to see if
there is heterogeneity in the association of \texttt{ct} with
\texttt{mort}{]}:

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = impact) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect group term  estimate std.error statistic    p.value conf.low conf.high
   <chr>  <chr> <chr>    <dbl>     <dbl>     <dbl>      <dbl>    <dbl>     <dbl>
 1 fixed  <NA>  (Int~    0.101   0.0173     -13.4   4.94e- 41   0.0723     0.141
 2 fixed  <NA>  type~    0.713   0.123       -1.96  5.01e-  2   0.509      1.00 
 3 fixed  <NA>  age      1.03    0.00165     20.2   2.13e- 90   1.03       1.04 
 4 fixed  <NA>  moto~    0.553   0.0386      -8.50  1.95e- 17   0.482      0.634
 5 fixed  <NA>  moto~    0.405   0.0289     -12.7   8.43e- 37   0.352      0.466
 6 fixed  <NA>  moto~    0.275   0.0202     -17.6   1.67e- 69   0.239      0.318
 7 fixed  <NA>  pupi~    3.73    0.230       21.4   2.09e-101   3.31       4.21 
 8 fixed  <NA>  pupi~    1.87    0.133        8.80  1.36e- 18   1.63       2.15 
 9 fixed  <NA>  ctIII    2.25    0.157       11.6   5.12e- 31   1.96       2.58 
10 fixed  <NA>  ctIV~    2.30    0.136       14.0   9.47e- 45   2.05       2.58 
11 ran_p~ name  sd__~    0.277  NA           NA    NA          NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

@@

This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
and \texttt{pupil}.{]} TODO: remove axis labels

@@

The first imputation model that we'll use is likely to be invalid. We do
\emph{not} use the cluster identifier \texttt{name} as imputation model
predictor. With this model, we ignore the multilevel structure of the
data, despite the high ICC. This assumes exchangeability between units.
We include it purely to illustrate the effects of ignoring the
clustering in our imputation effort. We'll use the default imputation
methods in \texttt{mice()} (predictive mean matching to impute the
continuous variables and logistic regression to impute binary
variables).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{center}

Updated until here!

\end{center}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

>>>>

A key characteristic of the missing data to take into account in
analyses is the mechanism behind the missingness. We distinguish between
MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice
this distinction is less clear. Since the essence of the true
non-response mechanism may not be known, it is generally inferred or
assumed to be ignorable (i.e., MCAR or MAR). {[}TODO: add that this
assumption may not always be valid, especially with modern types of big
data sources.{]} Depending on the actual missingness-generating
mechanism, missing data handling strategies may be more or less
suitable, see e.g., \citet{yuce08} and \citet{hox15}.

@@

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset'').{]} {[}TODO: Add that the
more the random effects are of interest, the more you need multilevel
imputation models.{]} {[}TODO: Add an overview of all possible predictor
matrix values in manuscript or \texttt{ggmice} legend.{]}

@@

{[}TODO: add where novice vs more experienced readers should start
reading.{]} For each of these datasets, we will discuss the nature of
the missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
\end{CodeInput}
\end{CodeChunk}

@@

This shows us that there are no apparent differences in the distribution
of \texttt{texp} depending on the missingness indicator of
\texttt{popular} (t = -0.873, p = 0.383). {[}TODO: think about what is a
meaningful rule of thumb to signal that the user should be worried?{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice(popmis, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- with(imp, 
+             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
\end{CodeInput}
\end{CodeChunk}

Print the estimates:

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\begin{CodeOutput}

@@

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     3.445     0.153    22.473    20.745     0.000     0.783     0.486 
sex             1.175     0.205     5.728    13.364     0.000     1.208     0.602 
texp            0.103     0.010     9.991    15.982     0.000     1.001     0.553 
sex:texp       -0.020     0.014    -1.460    11.610     0.171     1.421     0.644

                            Estimate 
Intercept~~Intercept|school    0.158 
sex~~sex|school                0.186 
Intercept~~sex|school         -0.025 
Residual~~Residual             0.704 
ICC|school                     0.183

@@

{[}TODO: Add that it doesn't work with systematic missingness (only with
sporadic). There's some pros and cons, and it may not even differ much
if the number of clusters is low.{]}

@@

\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp <- mice(popmis, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- with(imp, 
+             lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
\end{CodeInput}
\end{CodeChunk}

Print the estimates:

\begin{CodeChunk}
\begin{CodeInput}
R> testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{CodeInput}
\begin{CodeOutput}

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     3.341     0.179    18.648    14.134     0.000     1.137     0.587 
sex             1.267     0.216     5.866    12.916     0.000     1.255     0.612 
texp            0.108     0.012     9.061    12.109     0.000     1.351     0.631 
sex:texp       -0.030     0.014    -2.058    11.313     0.063     1.467     0.651

                            Estimate 
Intercept~~Intercept|school    0.198 
sex~~sex|school                0.231 
Intercept~~sex|school         -0.042 
Residual~~Residual             0.622 
ICC|school                     0.242

@@

The \texttt{impact} data included in the \pkg{metamisc} package is a
complete data set. The original data has already been imputed once
(Steyerberg et al, 2008). For the purpose of this tutorial we have
induced missingness (mimicking the missing data in the original data set
before imputation). The resulting incomplete data can be accessed from
\href{https://zenodo.com}{zenodo link to be created}.

@@

The estimated effects in the complete data are visualized in Figure
\ref{}.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # fit <- glmer(mod, family = "binomial", data = impact) # fit the model
R> # tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
\end{CodeInput}
\end{CodeChunk}

@@

This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
and \texttt{pupil}.{]}

@@
